\documentclass[11pt]{article}

\usepackage[a4paper,fancysections,titlepage]{src/polytechnique}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, fit, backgrounds, shapes.geometric}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{setspace}
\usepackage[patch=none]{microtype}

\setlength{\parskip}{0.7em}
\setstretch{1.05}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{bleu303pale},
  rulecolor=\color{bleu303},
  keywordstyle=\color{bleu315}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{rouge485},
  language=Python,
  numbers=left,
  numberstyle=\tiny\color{gray},
  xleftmargin=1.5em,
  framexleftmargin=1.5em,
}

\newcommand{\code}[1]{\mbox{\ttfamily\detokenize{#1}}}

\author{Nathan Duboisset}
\date{2026}
\title{PlaylistGenerator}
\subtitle{Semantic Music Retrieval from Natural Language and Visual Prompts}

\begin{document}

\maketitle

\tableofcontents

\newpage

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
% ─────────────────────────────────────────────────────────────────────────────

Music streaming platforms host hundreds of millions of tracks, yet their recommendation
engines still rely largely on collaborative filtering and rigid keyword tags.
Our \textbf{PlaylistGenerator} addresses this by building a semantic retrieval system that
maps audio, text and visual queries into a shared embedding space using CLAP~\cite{elizalde2023clap}. Beyond retrieval, the system generates contextual visual content to provide a complete user experience.

The system is built around five components:
\begin{enumerate}
    \item \textbf{Audio indexing} --- We pre-compute CLAP audio embeddings for every track
          in FMA Small and store them in a compressed \texttt{.npz} index.
    \item \textbf{Text retrieval} --- We encode the user prompt with the CLAP text encoder
          and run a cosine $k$-NN search against the index.
    \item \textbf{Two-phase fine-tuning} --- We (i) fine-tune the text encoder with LoRA on
          MusicCaps to enrich the textual embeddings for audio retrieval; (ii) further fine-tune the last stage of the audio encoder
          end-to-end on raw waveforms to tighten genre clusters in audio embedding space.
    \item \textbf{Image-to-playlist bridge} --- a small MLP projects CLIP image features
          into the CLAP text space to enable image-conditioned queries.
    \item \textbf{Visual Generation} --- We utilize Stable Diffusion to generate high-quality playlist covers based on the synthesized text prompt.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Background}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{CLAP}

CLAP~\cite{elizalde2023clap} is a dual-encoder model: the audio branch is an
\textbf{HTSAT-base} hierarchical Swin transformer (4 stages); the text branch is
\textbf{RoBERTa}. Both encode into a shared 512-dimensional space trained with the
symmetric InfoNCE loss:
\begin{equation}
    \mathcal{L} = \frac{1}{2}\Bigl(
        \mathcal{L}_{a\to t} + \mathcal{L}_{t\to a}
    \Bigr), \quad
    \mathcal{L}_{a\to t} = -\frac{1}{N}\sum_{i=1}^{N}
        \log\frac{e^{s\,\mathbf{a}_i^\top\mathbf{t}_i}}
                  {\sum_{j} e^{s\,\mathbf{a}_i^\top\mathbf{t}_j}}
    \label{eq:infonce}
\end{equation}
where $s = \exp(\texttt{logit\_scale})$ is a learnable temperature parameter and $N$ is the batch size.
We used the \texttt{music\_audioset\_epoch\_15\_esc\_90.14.pt} checkpoint which achieves 90.14 \% accuracy on the ESC-50 benchmark.

\subsection{LoRA and CLIP}

To adapt the model to the specific nuances of musical descriptions, we employ LoRA~\cite{hu2022lora} for parameter efficient fine-tuning the layers of RoBERTa, the text encoder(see Section~\ref{sec:phase1}).

For visual modality, we leverage CLIP (ViT-L/14) to produce 1024-dimensional image embeddings. These embeddings serve as the input for our bridge MLP, which aligns visual semantics with the music-text latent space (Section~\ref{sec:bridge}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Datasets}
% ─────────────────────────────────────────────────────────────────────────────
To develop and evaluate the PlaylistGenerator, we leverage three distinct datasets for audio indexing, semantic alignment, and cross-modal bridge training.\begin{itemize}
\item \textbf{FMA Small (Audio Indexing)}~\cite{defferrard2017fma}: This dataset provides the retrieval library, consisting of 8,000 30-second tracks across eight genres. After excluding six corrupted files, 7,994 tracks were indexed for $k$-NN search.
\item \textbf{MusicCaps (Text-Audio Fine-tuning)}~\cite{agostinelli2023musiclm}: Utilized for Phase 1 LoRA adaptation, this dataset contains 5,521 audio clips captioned by expert musicians. After processing, 4,357 valid samples were paired with audio embeddings. These are split into training (3,704) and testing (653) sets (85/15 ratio). For Phase 2, the model accesses the raw WAV files directly to specialize the audio encoder.

\item \textbf{MS COCO Captions (Cross-modal Alignment)}: To train the CLIP-CLAP bridge MLP, we used 20,000 image--caption pairs from the MS COCO dataset. Visual features are extracted via a frozen CLIP ViT-L/14, while text features are processed by the frozen CLAP text encoder to align both modalities.

\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Methods}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Overview}
\begin{figure}[h]
\centering
% \resizebox{\textwidth}{!}{ % Décommente cette ligne (et l'accolade de fin) si le schéma dépasse dans les marges
\begin{tikzpicture}[
    font=\small,
    box/.style={draw, rounded corners=3pt, minimum width=2.6cm, minimum height=0.8cm,
                align=center, fill=blue!10, draw=blue!50!black},
    model/.style={draw, rounded corners=3pt, minimum width=2.6cm, minimum height=0.8cm,
                  align=center, fill=red!10, draw=red!70!black},
    % Le bug de la zone grise est corrigé ici en retirant "cylinder"
    store/.style={draw, rounded corners=3pt, minimum width=2.6cm, minimum height=0.8cm,
                  align=center, fill=gray!10, draw=gray!70!black},
    arr/.style={-{Stealth[length=5pt]}, thick},
    node distance=1.2cm
]

% --- NODES LAYOUT ---

% Text Path (Niveau Y = 5)
\node[box]   (textprompt)  at (0, 5)    {Text Prompt};
\node[model] (claptext)    at (3.5, 5)  {CLAP Text Enc.\\(RoBERTa + LoRA)};

% Image Path (Niveau Y = 2.5)
\node[box]   (imageprompt) at (0, 2.5)  {Image Prompt};
\node[model] (clip)        at (3.5, 2.5){CLIP ViT-L/14\\Visual Encoder};
\node[model] (bridge)      at (7, 2.5)  {Bridge MLP\\$1024 \to 512$};

% Audio Path (Niveau Y = 0)
\node[box, fill=red!10, draw=red!50!black]  (audiocorpus) at (0, 0)    {FMA Small};
\node[model] (clapaudio)   at (3.5, 0)  {CLAP Audio Enc.\\(HTSAT)};
\node[store] (index)       at (7, 0)    {\texttt{.npz} Index\\$7994 \times 512$\\offline embeddings};

% Generation Path (Niveau Y = 6.5)
\node[model] (sd)          at (7, 6.5)  {Stable Diffusion\\(segmind/small-sd)};

% --- STOCHASTICITY (NOISE) --- (Niveau Y = 3.75, pile au milieu)
\node[font=\small, text=black!80] (noise) at (7, 3.75) {Noise $\epsilon$};
\node[draw, circle, inner sep=1.5pt, fill=yellow!20, draw=yellow!60!black] (sum) at (9, 3.75) {$\oplus$};

% --- PROCESSING & OUTPUTS ---
\node[model] (knn)         at (11.5, 3.75){Cosine $k$-NN\\Search};
\node[box, fill=green!10, draw=green!50!black] (playlist) at (14.5, 3.75) {\textbf{Playlist}\\(Audio Tracks)};
\node[box, fill=green!10, draw=green!50!black] (cover)  at (14.5, 6.5)  {\textbf{Cover Art}\\(Image)};

% --- ARROWS ---

% Text Flow vers k-NN
\draw[arr] (textprompt) -- (claptext);
\draw[arr] (claptext.east) -- node[above]{$e_t$} (9, 5) -- (sum.north);

% Text Flow vers Stable Diffusion
\draw[arr] (textprompt.north) |- (sd.west);
\draw[arr] (sd.east) -- (cover.west);

% Image Flow
\draw[arr] (imageprompt) -- (clip);
\draw[arr] (clip) -- (bridge);
\draw[arr] (bridge.east) -- node[above]{$e_v$} (9, 2.5) -- (sum.south);

% Noise Flow (L'addition)
\draw[arr] (noise.east) -- (sum.west);
\draw[arr] (sum.east) -- node[above]{$e_q$} (knn.west);

% Audio Flow (L'index passe par le bas pour ne rien croiser)
\draw[arr] (audiocorpus) -- (clapaudio);
\draw[arr] (clapaudio) -- (index);
\draw[arr] (index.east) -- node[above, pos=0.4]{} (11.5, 0) -- (knn.south);

% Final Output
\draw[arr] (knn.east) -- (playlist.west);

\end{tikzpicture}
% } % Décommente cette accolade si tu as utilisé le \resizebox
\caption{Multimodal architecture of the system. A stochastic noise $\epsilon$ is added to the query embedding $e_q$ prior to the $k$-NN search to ensure diverse playlist generation.}
\label{fig:arch}
\end{figure}


\subsection{Phase 1 — LoRA Text Encoder Fine-tuning}
\label{sec:phase1}

The text encoder (RoBERTa) is adapted to music-specific captions via LoRA
(\texttt{r=16}, $\alpha=32$, dropout=0.1, targets: \texttt{query}, \texttt{value}),
adding \textbf{1,246,209 trainable parameters} (0.62\% of 200M). The audio encoder
is \textbf{fully frozen}; audio embeddings are pre-computed once and cached.

Each training step minimises the symmetric InfoNCE loss (Eq.~\ref{eq:infonce}) between
a batch of cached audio embeddings and freshly encoded text embeddings. Optimization is performed using AdamW with a learning rate of $10^{-4}$, a weight decay of $10^{-4}$, and a cosine annealing schedule over 3 epochs. We use a batch size of 16 and apply gradient clipping at a threshold of 1.0 to ensure numerical stability.

\subsection{Phase 2 — Audio Encoder Fine-tuning}
\label{sec:phase2}

In Phase 2, we shift our focus toward specializing the audio modality. While the text encoder remains frozen with its LoRA weights from Phase 1, we selectively unfreeze the \textbf{last Swin stage}
(\texttt{audio\_branch.layers[3]}), the final norm,
\texttt{audio\_transform}, and \texttt{audio\_projection}. This targeted strategy, affecting approximately the top 15\% of the encoder's depth, preserves
low-level spectral features while specialising high-level representations to music.

Because the audio encoder is now updated, pre-computed embeddings are stale.
Training instead loads raw 10-second WAV files, resampled on-the-fly to 48\,kHz
(CLAP's native rate), and passes them through the live audio encoder. The text
encoder runs in eval mode (no LoRA dropout) with no gradient. The loss is the same
InfoNCE objective; the LoRA text embeddings serve as fixed targets.

Optimization is performed using AdamW with a learning rate of $lr=5\times10^{-6}$ (conservative to avoid forgetting), a weight decay of $10^{-4}$, and a cosine learning rate over 2 epochs, a batch size of 8, and gradient clipping at 1.0 epochs.


\subsection{\texorpdfstring{CLIP$\to$CLAP}{CLIP-to-CLAP} Bridge MLP}
\label{sec:bridge}

To enable image-conditioned playlist generation without using a VLM to convert the image input into a textual description, we implement a cross-modal bridge designed to align visual embeddings with those from the audio and text encoders from CLAP. This module acts as a translator, mapping the 1024-dimensional image embeddings produced by a frozen CLIP ViT-L/14 encoder onto the 512-dimensional space of the CLAP model.

The bridge is structured as a Multi-Layer Perceptron (MLP) with two hidden layers ($1024 \to 1024 \to 1024 \to 512$) and ReLU activation function, Layer Normalization, and a dropout rate of $0.2$.

We train this MLP using a one-directional InfoNCE loss ($\tau=0.07$) on a dataset of 20,000 image--caption pairs from MS COCO. During the training procedure, both the CLIP visual encoder and the CLAP text encoder are kept frozen; the MLP learns to minimize the distance between the projected image embedding and the corresponding CLAP text embedding. The model is optimized using AdamW with a learning rate of $2 \times 10^{-4}$ and a batch size of 64 over 5 epochs.


\subsection{Visual Synthesis}
\label{sec:visual_gen}

To complement the semantic retrieval of audio tracks and offer a complete user experience, we add a generative vision component designed to produce context-aware playlist covers (when the input is not already an image).  We utilize the \texttt{segmind/small-sd} model, a distilled version of Stable Diffusion. This distilled model enables computational efficiency and low inference time, while maintaining high-fidelity synthesis.

The generation is conditioned by the textual input used for the initial audio retrieval, which we encapuslate into a specific prompt for the generated images to resemble playlist covers :
\textit{"Album cover art for a music playlist: [prompt], vibrant colors, artistic, detailed, professional album artwork, 4k."}, while a negative prompt is also given to produce more satisfying results : \textit{"text, watermark, blurry, low quality, distorted"}.

The synthesis process is configured with 30 inference steps and a guidance scale of 7.5, to achieve a balance between adherence to the prompt and artistic creativity. The resulting image is created in 512$\times$512 resolution.


\subsection{Evaluation Metrics}

To quantify the impact of our two-phase fine-tuning strategy, we employ two complementary metrics. The first evaluates the semantic bridge between modalities (Phase 1), while the second assesses the structural density of the audio latent space itself (Phase 2).

\paragraph{Text--audio alignment.} This metric measures the success of the semantic adaptation performed in Phase 1. It is defined as the mean cosine similarity between the CLAP text embedding of a genre name (e.g., "Electronic") and the centroid of the audio embeddings for all tracks belonging to that genre within the FMA Small dataset. This alignment score measures how effectively the text encoder retrieves the relevant audio clusters in the shared latent space.

\paragraph{Intra-genre cohesion.} This metric assesses the refinement of the audio manifold during Phase 2. It is calculated as the mean \emph{pairwise} cosine similarity among all audio embeddings within a specific genre:
\[
    C_g = \frac{1}{n_g(n_g-1)}\sum_{i \neq j} \mathbf{a}_i^\top \mathbf{a}_j
\]
A higher $C_g$ means tracks of the same genre are more tightly clustered together in the embedding space, measuring directly how discriminative and specialized the audio representations are.
This metric is unaffected by the text-only updates of Phase 1, but serves as the primary indicator of the audio encoder’s specialization in Phase 2.



% ─────────────────────────────────────────────────────────────────────────────
\section{Experiments and Results}
% ─────────────────────────────────────────────────────────────────────────────

In this section, we evaluate the impact of our two-phase fine-tuning on semantic alignment and genre clustering. We first present the quantitative improvements brought by Phase 1, followed by a deeper analysis of the audio manifold in Phase 2.

\subsection{Phase 1 — LoRA Text Encoder}
The objective of Phase 1 was to adapt the generic CLAP text encoder to the domain-specific vocabulary of music descriptions. Table~\ref{tab:training} tracks the evolution of the symmetric InfoNCE loss. We observe a steady convergence over 3 epochs.

\begin{table}[h]
\centering

\label{tab:training}
\begin{tabular}{ccc}
\toprule
\textbf{Epoch} & \textbf{Train} & \textbf{Val} \\
\midrule
1 & 0.6165 & 0.4093 \\
2 & 0.3959 & 0.3659 \\
3 & 0.3277 & 0.3428 \\
\bottomrule
\end{tabular}
\caption{InfoNCE loss per epoch — Phase~1 (MusicCaps, 3704/653 samples).}
\end{table}

\begin{table}[h]
\centering
\label{tab:cosine}
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Avg.\ cosine sim.} & \textbf{Gain} \\
\midrule
Baseline CLAP         & 0.3135 & --- \\
CLAP + LoRA (Phase~1) & 0.3710 & $+18.3\%$ \\
\bottomrule
\end{tabular}
\caption{Average cosine similarity on the MusicCaps test set (653 samples).}
\end{table}

LoRA leads to an absolute gain of \textbf{+0.0575} in cosine similarity, representing a \textbf{+18.3\% improvement} over the baseline, although only 0.62\%
of the parameters are updated.

\subsection{Per-genre Text--Audio Alignment}

Table~\ref{tab:genre_align} shows per-genre alignment before and after Phase~1.
Genres with low baseline alignment (Electronic, Experimental) see the largest gains;
genres already well-aligned (Folk, Hip-Hop) regress slightly, likely because
MusicCaps skews toward Electronic/Experimental styles.

\begin{table}[h]
\centering
\label{tab:genre_align}
\begin{tabular}{lccc}
\toprule
\textbf{Genre} & \textbf{Baseline} & \textbf{Phase 1} & \textbf{$\Delta$} \\
\midrule
Electronic    & 0.1085 & 0.2041 & $+0.0956$ \\
Experimental  & 0.2349 & 0.2856 & $+0.0507$ \\
Instrumental  & 0.2466 & 0.2706 & $+0.0240$ \\
Rock          & 0.3540 & 0.3271 & $-0.0269$ \\
Hip-Hop       & 0.4212 & 0.3471 & $-0.0741$ \\
Folk          & 0.4135 & 0.3133 & $-0.1001$ \\
Pop           & 0.2537 & 0.1383 & $-0.1155$ \\
International & 0.3483 & 0.1850 & $-0.1632$ \\
\bottomrule
\end{tabular}
\caption{Per-genre text--audio cosine similarity on FMA Small.}
\end{table}

\subsection{Intra-genre Cohesion and Phase 2}

Table~\ref{tab:cohesion} reports the intra-genre cohesion $C_g$ before Phase~2
(frozen audio encoder) and after Phase~2 (fine-tuned last Swin stage). Because
Phase~1 does not touch the audio encoder, the ``Baseline'' and ``Phase~1'' columns
are identical; only Phase~2 changes $C_g$.

\begin{table}[h]
\centering
\label{tab:cohesion}
\begin{tabular}{lcc}
\toprule
\textbf{Genre} & \textbf{Phase 1 (frozen audio)} & \textbf{Phase 2 ($\Delta$)} \\
\midrule
Hip-Hop       & \emph{run notebook} & --- \\
Folk          & \emph{run notebook} & --- \\
Rock          & \emph{run notebook} & --- \\
Electronic    & \emph{run notebook} & --- \\
Experimental  & \emph{run notebook} & --- \\
Instrumental  & \emph{run notebook} & --- \\
Pop           & \emph{run notebook} & --- \\
International & \emph{run notebook} & --- \\
\bottomrule
\end{tabular}
\caption{Intra-genre audio cohesion $C_g$ on FMA Small.}
\end{table}

An increase in $C_g$ after Phase~2 indicates that the fine-tuned audio encoder
maps same-genre tracks closer together, making $k$-NN retrieval more genre-coherent.

\subsection{Qualitative Retrieval}

Table~\ref{tab:qualitative} shows top-5 MusicCaps results for three prompts with
the Phase~1 model. Retrieved captions are semantically consistent with the query
in all three cases.

\begin{table}[h]
\centering
\label{tab:qualitative}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{clc}
\toprule
\textbf{Rank} & \textbf{Caption (truncated)} & \textbf{Score} \\
\midrule
\multicolumn{3}{l}{\textit{``slow chill hip hop with a mellow bassline''}} \\
\midrule
1 & Hip hop song with consistent low-quality bassline   & 0.405 \\
2 & Dark moody piano melody with hip hop beat           & 0.392 \\
3 & Instrumental movie soundtrack theme                 & 0.390 \\
\midrule
\multicolumn{3}{l}{\textit{``upbeat reggae with rhythmic percussion''}} \\
\midrule
1 & Acoustic drum playing a reggae groove with rim shot & 0.421 \\
2 & Female singer over a cool melody                    & 0.413 \\
3 & Reggae song with flat male vocal rapping            & 0.384 \\
\midrule
\multicolumn{3}{l}{\textit{``sad orchestral strings and piano''}} \\
\midrule
1 & Sustained strings melody, low quality recording     & 0.407 \\
2 & Classical wide mellow strings melody                & 0.402 \\
3 & Mellow piano melody, low quality recording          & 0.399 \\
\bottomrule
\end{tabular}
\caption{Top-5 retrieval from MusicCaps (Phase~1 model).}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
% ─────────────────────────────────────────────────────────────────────────────
In this project, we developed a comprehensive multimodal system for semantic music retrieval and visual synthesis. We leveraged the CLAP architecture to bridge the gap between natural language/visual queries and raw audio signals.

Our two-phase fine-tuning strategy demonstrated the efficiency of targeted parameter updates. We first utilized LoRA to adapt the RoBERTa text encoder to specialized musical captions, yielding a +18.3\% improvement in text--audio cosine similarity with only 0.62\% of parameters trained. We then focused on the audio encoder, unfreezing the final Swin transformer stages to refine intra-genre cohesion on raw waveforms.

Beyond retrieval, we designed and trained a CLIP-to-CLAP bridge MLP to enable and a Stable Diffusion generative pipeline completes the tool into an end-to-end playlist generator, a feature which is not already available on streaming platforms.

Future work could explore the integration of Active User Feedback (RLHF) using triplet loss to further align the latent space with individual preferences.
% ─────────────────────────────────────────────────────────────────────────────
\begin{thebibliography}{9}

\bibitem{elizalde2023clap}
B.~Elizalde, S.~Deshmukh, M.~Al~Ismail, H.~Wang.
\textit{CLAP: Learning Audio Concepts from Natural Language Supervision}.
ICASSP 2023.

\bibitem{radford2021clip}
A.~Radford, J.~W.~Kim, C.~Hallacy \emph{et al.}
\textit{Learning Transferable Visual Models From Natural Language Supervision}.
ICML 2021.

\bibitem{hu2022lora}
E.~J.~Hu, Y.~Shen, P.~Wallis \emph{et al.}
\textit{LoRA: Low-Rank Adaptation of Large Language Models}.
ICLR 2022.

\bibitem{defferrard2017fma}
M.~Defferrard, K.~Benzi, P.~Vandergheynst, X.~Bresson.
\textit{FMA: A Dataset for Music Analysis}.
ISMIR 2017.

\bibitem{agostinelli2023musiclm}
A.~Agostinelli, T.~I.~Denk, Z.~Borsos \emph{et al.}
\textit{MusicLM: Generating Music From Text}.
arXiv:2301.11325, 2023.

\end{thebibliography}

\end{document}
